---
title: "Descriptive Statistics in R"
author: "Victor A. Ordu"
date: '2021-07-16'
output:
  ioslides_presentation: default
  beamer_presentation: default
---

# Introduction

------------------------------------------------------------------------

-   Most of the functions related to pure statistics can be found in the *base* and *stats* package
-   They are always pre-loaded by default in every R session

```{r}
search()
```

# Summarisation of Data

## Summation

-   Numerical data

The sum:

$$Sum(X) = \sum_{i=1}^{N}x_i$$ where $x_i$ is a single element in a numeric vector and $N$ is the number of elements in that vector.

We add up the values with `sum()`:

```{r}
x <- 1:100
sum(x)
```

------------------------------------------------------------------------

-   Categorical data

Here we do **counting** with `length()`

How many letters are there in the English alphabet?

```{r}
length(LETTERS)
```

## Scaling

-   Addition/subtraction
-   Multiplication and/or division.
-   When you apply a constant number, it computes on every element of the vector

```{r, results='markup'}
mix <- c(5, 0, 28, -5, 29, 77, 10, 57, 28, 88, 298)

mix + 5
mix * 2
mix / 3
mix %% 2
```

## Maxima and minima

-   Get the highest or lowest value in a collection.
-   Functions: `max()` and `min()`

```{r}
max(mix)
min(mix)
```

## Summarization

-   Tukey's five-number summary: minimum, maximum, mean, 1st quartile, 3rd quartile
-   At a glance, gives you an idea of central statistic, dispersion and extremes of the data.

```{r}
fivenum(mix)
```

## Sorting

Create an ordered array of values

```{r}

sort(mix)
```

In descending order

```{r}
sort(mix, decreasing = TRUE)
```

You can sort character vectors as well

```{r}
sort(c("apple", "grapple", "tuple", "hubble", "ample"))
```

## Ranking

Get the rank of values within a collection

```{r}
order(mix)
```

Compare with original vector for clarity:

```{r}
comp <- data.frame(original = mix, ordered = order(mix))
comp
```

# Measures of Central Tendency

## Arithmetic mean

Arithmetic mean, of a population $\mu$ is defined by the equation $$\mu = \frac{\sum_{i=1}^Nx_i}{N}$$ - Computed with the function `mean()`

```{r}
mean(mix)
```

## Median

The median, the *middle* value in an ordered array, $X$, with $N$ numbers can be described by the formula $$median(X) = x_{(N+1)/2}$$ and when considering even numbered sets, we have the formula $$median(X)=\frac{x_{N/2}+x_{(N/2)-1}}{2}$$ ---

-   Computed with the function `median()`

```{r}
median(mix)
```

## The Difference between mean and median

```{r}
# Add an outlier to the vector 'mix'
mix[length(mix) + 1] <- 1000
mix
mean(mix)
median(mix)
```

## Mode

There is no function per se in R for computing the mode, but developing a custom function for this purpose is trivial and would be done on a case-by-case basis.

# Measures of Dispersion

------------------------------------------------------------------------

When you want to assess the variability in a set of data or a given variable

## Range

The range, $R$, is the difference between the lowest value, $x_S$ and the largest value, $x_L$, i.e. $$R=x_L-x_S$$ In R, the function `range` returns both values as a 2-element numeric vector.

```{r}
range(mix)
```

## Interquartile Range

-   Is the range between the first and the third quartile
-   A measure of the "middle fifty percent" of a dataset
-   It is the range between the 75th and 25th percentile

$$IQR=Q_3-Q_1$$

```{r}
IQR(mix)
```

------------------------------------------------------------------------

In the box-and-whiskers plot, the "box" displays the $IQR$.

```{r}
boxplot(mix)
```

## Variance

Dispersion around the mean, with the difference squared to create absolute values.

$$Var(X)=\frac{\sum_{i=1}^{N}(x_i-\mu)^2}{N}$$

```{r}
var(mix)
```

## Standard Deviation

This is essentially (elementarily) the square root of the variance

$$s.d.(X)=\sqrt{\frac{\sum_{i=1}^{N}(x_i-{\bar{x}})^2}{N}}$$

```{r}
sd(mix)
```

## Coefficient of Variation

Used to measure the variation between 2 different but related datasets It is the ratio between the standard deviation and the mean, expressed as a percentage.

$$C.V.=\frac{s.d.}{\bar{x}}(100)$$ \# Grouping of Data

## Numerical vs. Categorical Data

-   With numerical data we create intervals first, coverting them to discrete categories.
-   Then we create a frequency tabulation

```{r}
cut(mix, 3)  # create 3 groups using defaults
```

------------------------------------------------------------------------

```{r}
cut(mix, c(-5, 0, 100, 1000))  # Note missing value
```

\`

------------------------------------------------------------------------

```{r}
cut(mix, c(-4, 0, 100, 1000), include.lowest = TRUE)
```

------------------------------------------------------------------------

```{r}
cut(mix, breaks = c(min(mix), median(mix), max(mix)))
```

## Frequency Distribution

-   For categorical/discrete data
-   In R, we use the `table()` function.

```{r}
table(esoph$agegp)
```

## Cumulative Frequency Distribution

```{r}
cumsum(mix)
```

------------------------------------------------------------------------

```{r}
# Compare with original
cumfrq <- data.frame(original = mix, cum.freq = cumsum(mix))
cumfrq
```

Something is wrong with this arrangement. Can you spot it?

# Some important characteristics of vectors

------------------------------------------------------------------------

-   We need to discuss some important behaviours of vectors
-   They have practical implications when you are carrying out analyses.

## Vectorization

-   A process whereby an operation carried out affects every element of a vector.
-   First let's create a vector with random whole numbers

```{r}
set.seed(134)
myvalues <- sample(1:7, size = 10, replace = TRUE)
myvalues
```

## Arithmetic

```{r}
myvalues + 1
```

------------------------------------------------------------------------

```{r}
myvalues * 5
```

------------------------------------------------------------------------

```{r}
myvalues < 3
```

## Vectorized Logical 'AND'

```{r}
set.seed(54)
newvalues <- sample(myvalues, size = length(myvalues), replace = TRUE)
identical(newvalues, myvalues)
```

------------------------------------------------------------------------

```{r}
myvalues < newvalues
```

```{r}
myvalues > 2 && myvalues < 5
```

------------------------------------------------------------------------

But with "Vectorized" `&` we do this check element by element.

```{r}
myvalues > 2 & myvalues < 5
```

------------------------------------------------------------------------

-   `&&` and `||` are best for logical single outcome computations e.g. `is.character(x) && is.atomic(x)`

-   `&` and `|` are used to compare elements one by one along the length of two vectors. e.g. `is.na(x) | duplicated(x)`

## Recycling

When 2 vectors are used together as operands in a computation e.g. $A + B$, if `length(A) != length(B)`, the shorter one will be recycled until the computation on the longer vector is completed, without prejudice to which element of the shorter is used last.

------------------------------------------------------------------------

Thus,

```{r}
x <- 5
y <- 1:8
x * y    # Recycling of lengh 1L akin to vectorization
```

------------------------------------------------------------------------

```{r}
x <- c(5, 10)
x * y
```

------------------------------------------------------------------------

Display side-by-side

```{r}
data.frame(y, prod = x * y)
```

------------------------------------------------------------------------

Demonstrate abrupt stop to recycling

```{r, warning=FALSE}
x[3] <- 15L
x * y
```

------------------------------------------------------------------------

Viewed in a data frame...

```{r, echo=FALSE, warning=FALSE}
data.frame(x = rep_len(x, 8), y, `x times y` = x * y)
```

## Missing values

Most computations, such as the ones above will return `NA` if a missing value is present.

```{r}
mix[7] <- NA
mix

sum(mix)
```

------------------------------------------------------------------------

To fix this, remove `NA` from the computation. See the `sum`'s signature:

```{r}
args(sum)
```

Where `...` received objects to be summed up and `na.rm` tunes our ability to discount `NA` from the computation.

------------------------------------------------------------------------

Thus,

```{r}
sum(mix, na.rm = TRUE)
```

------------------------------------------------------------------------

Many other functions used for statistical computations have arguments for dealing with `NA`. Whether you leave or remove depends entirely on the circumstance at any given time.

```{r}
mean(mix)
```

------------------------------------------------------------------------

```{r}
mean(mix, na.rm = TRUE)
```

------------------------------------------------------------------------

The real point is to always glance at the help files e.g. `?mean`. The funcion `args()` used earlier also provides a way to quickly look at a function signature, but it is not always guaranteed to be informative, especially with generic functions.

```{r, eval=FALSE}
args(plot)

?plot
```

## Peculiarities of Logical Vectors

-   Recall: `0` evaluates to `TRUE` and any other number evaluates to `FALSE`.
-   Thus:

```{r}
as.logical(0)
as.logical(45)
as.logical(-66)
as.logical(0.004)
```

------------------------------------------------------------------------

```{r}
as.logical(Inf)
as.logical(-Inf)
```

Missing values are logical vectors but do not return logical values.

```{r}
is.logical(NA)
typeof(NA)
as.logical(NA)
```

## Summarising Logical Vectors

-   Technically, within a vector, `TRUE` will be taken as `1`, while `FALSE` is returned as `0`.
-   This is very useful for summarising data

```{r}
set.seed(45)
logi <- sample(c(T, F), 1e6, replace = TRUE, prob = c(.3, .7))
sum(logi)
```

This compares well when we use 1s and 0s instead:

```{r}
set.seed(45)
numi <- sample(c(1, 0), 1e6, replace = TRUE, prob = c(.3, .7))
sum(numi)
```

------------------------------------------------------------------------

-   In summarising logical vectors, we can easily use the same construct to compute the mean.

-   But first, a digression:

    -   To proceed, we have retype or copy-paste, violating the [**DRY principle**](https://en.wikipedia.org/wiki/Don%27t_repeat_yourself)

    -   Write a function that generates a binary vector of length $10^6$

    -   Perform some input validation to ensure that only a vector of length `2L` is used

    -   Check that the input satisfies the following conditions:

        -   If logical, MUST contain only `TRUE` and `FALSE`.
        -   If numeric, MUST be integers `0L` and `1L`.
        -   If strings, MUST be unique.

------------------------------------------------------------------------

We can easily calculate proportions i.e. percentages by calling `mean` on logical vectors.

```{r}
set.seed(45)
logi <- sample(c(T, F), 1e6, replace = TRUE, prob = c(.3, .7))
mean(logi)
```

## Logical subsetting

-   A logical value of `TRUE` will select an element in a vector, while `FALSE` will drop it.
-   Thus

```{r}
below30 <- mix < 30
below30
```

We can now subset this `mix` to get out the values that satisfy this condition:

```{r}
mix[below30]
```

The rules of recycling still apply here

```{r}
recyc <- mix[c(T, F, T)]
recyc
```

For a better view

```{r}
index <- rep_len(c(T, F, T), length(mix))
data.frame(mix = mix, index = index)
```

Using the function `which` one can discover the indices of a logical vector that evaluate to `TRUE`

```{r}
dat <- data.frame(
  name = c("Tom", "Dick", "Harry", "Sally"),
  sex = c("M", "M", "M", "F"),
  age = c(45, 29, 12, 23)
)
```

How many are older than 25?

```{r}
`25 plus` <- dat$age > 25
`25 plus`
```

But, **who** is older than 25?

```{r}
which(dat$age > 25)
```

Putting it all together we can write code in various ways to extract information from a data frame:

-   Who is older than 25?

```{r}
i <- which(dat$age > 25)
dat$name[i]
```

-   What is the gender of those older than 25?

```{r}
dat$sex[i]
```

-   What are the records of those older than 25?

```{r}
dat[i, ]
```

Note that creating the variable `i` is an optional step, depending on the coding situation and the choice on what to do depends on the balance between readability vs. convenience. The following gives the same result

```{r}
# Name of subjects older than 25
dat$name[dat$age > 25]
```

------------------------------------------------------------------------

Finally, we can use multiple comparisons in extracting data from a table. We will use the `esoph` dataset

Show records of women in the data that are younger than 25:

```{r}
dat[dat$age < 25 & dat$sex == "F", ]
```

# Homework

## Assignment 1

We said that the Coefficient of Variation "is the ratio between the standard deviation and the mean, expressed as a percentage".

**Task**: Create this function.

------------------------------------------------------------------------

**Your function should**:

-   Receive only a vector of numbers as input and throw an error that says "Illegal input" when this is violated.

-   Should work only for vectors with at least 2 elements and at the most 100 elements. If any of these are violated, throw an error that says "The vector is either too short or too long".

-   In the script write first the definition of your function and then a few test cases below.

## Assignment 2

Write R code to explore and describe the `esoph` dataset as follows: - Number of rows and columns - Total number of controls and total number of cases or esophageal cancer. - Display the categories of the age groups (Tip: Use the function `levels`) - Average number of cases aged 35 to 44 years. - How many of the aforementioned both used tobacco and drank alcohol?

## Submission

Post your code as a GitHub Gist. Then copy its link and post in the \*training\* on Slack.
