---
title: "Simple Plotting in R"
author: "Victor Ordu"
date: "26 June 2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- ## Outline -->

<!-- - Base graphics  -->
<!--     * Most of the functions are in the `graphics` package. -->
<!-- - `ggplot2` -->
<!--     * An exceptionally great package authored by Hadley Wickham. -->


## Our Dataset
>- We will work with the in-built `airquality` dataset.
>- It represents daily air quality measurements in New York from 1st May to 30th September 1973.
```{r data}
df <- airquality
str(df)
```

## Plotting of Numerical Data
>- Histograms are for representing **discrete** numerical data i.e. integers.
>- They are particularly good at displaying the 'skew' of the distribution around an estimator such as the arithmetic mean or the median.

*** 

*For Ozone:*
```{r}
hist(df$Ozone)
```

***

*For Solar Radiation:*
```{r hist2}
## Compare this with the other histogram
hist(df$Solar.R)
```

## Frequency Polygons
 >- Also good for discrete values.
 >- The function `freqpolygon()` is from the `mosaic` package. 

***

*For Ozone:*
```{r, mosaic, message=FALSE}
library(mosaic)
freqpolygon(df$Ozone)
```

*** 

*For Solar Radiation:*
```{r freqpoly2}
freqpolygon(df$Solar.R)
```

## Scatter Plots
>- Unlike the previous charts, which represent data as 'collectives', scatter plots are used to represent individual observations.
>- Quite useful in bivariate analyses - to examine relationship between 2 discrete numerical variables (e.g. correlation).

***

*For Ozone:*
```{r scatter}
## Not all that useful for univariate analysis
plot(df$Ozone)
```

***

*For Solar Radiation:*
```{r scatter2}
plot(df$Solar.R)
```

## Matrix of Scatter Plots

```{r pairs}
pairs(df)
```

## Do some data transformation
Create a **brand new** column called `Mth.Name`

- that contains the months May to September
- properly spelt out (rather than as numbers as we have in the original dataset).
- placed in chronological order (i.e. an ordinal variable).
```{r transform}
df$Mth.Name <- month.name[df$Month]
df$Mth.Name <-
  factor(df$Mth.Name, levels = month.name[5:9], ordered = TRUE)
```

***
View the results:
```{r month head}
head(df$Mth.Name)
```

```{r month tail}
tail(df$Mth.Name)
```

- We have created a new column called `Mth.Name` and made it an ordered factor (categorical variable) of the months of May through September.
- Can be confirmed using functions like `str()`, `dim()`, etc.

## Bar Charts

```{r bar chart}
tab <- table(df$Mth.Name)
barplot(tab)
```


## Box-and-Whiskers Plot
*For Ozone*:
```{r box-and-whiskers}
boxplot(df$Ozone)
```

***

*For Solar Radiation*:
```{r box-and-whiskers2}
boxplot(df$Solar.R)
```

## 1-D Scatter Plot
The one-dimensional scatter plot is useful for displaying 'spread' in small datasets.  

***

*For Ozone*:
```{r strip 1-D scatter}
stripchart(df$Ozone)
```

***

*For Solar Radiation*:
```{r}
stripchart(df$Solar.R)
```

<!-- ## Functions (*A Digression*) -->
<!-- Before we proceed, this is a good opportunity to introduce the use of functions in R. -->
<!-- We use functions to modularise our code i.e. encapsulate an expression or a series of expressions. Typically, functions are given a name and can be reused once they are loaded into an R session.  -->

<!-- Example 1 - we will create a function that returns the string `Hello, R!`. This function has no parameters. -->
<!-- ```{r} -->
<!-- ## Define a function -->
<!-- sayHi <- function() "Hello, R!" -->

<!-- ## Run the function -->
<!-- sayHi() -->
<!-- ``` -->

<!-- *** -->

<!-- Example 2 - we will create a function called `addTwo`, which will add `2` to any number that is passed to it as an argument -->
<!-- ```{r} -->
<!-- ## Define another function -->
<!-- addTwo <- function(x) { -->
<!--   x + 2 -->
<!-- } -->

<!-- ## Create a variable -->
<!-- num <- 42 -->

<!-- ## Run the function -->
<!-- addTwo(num)         # ditto addTwo(42) -->
<!-- ``` -->

## Time Series
<!-- - We have opted to create a function `my_ts` so that we can repeatedly run some expressions for plotting the time series. -->
<!-- - The function has a parameter `var` -->
<!-- - It will: -->
<!-- 1. Build a time-series object (using the `ts()` function) and -->
<!-- 2. Draw a plot. -->
<!-- ```{r define function} -->
<!-- my_ts <- function(var) { -->
<!--   ## Expression 1: Create a May-Sept time-series object internally -->
<!--   t <- ts(var, start = 5, end = 9, frequency = 5) -->

<!--   ## Expression 2: Display a plot of the time-series -->
<!--   plot(t) -->
<!-- } -->
<!-- ``` -->

<!-- ***  -->
<!-- We now run the function using the *Ozone* column as an argument to `var` -->
```{r run my_ts}
var = df$Ozone
t <- ts(var, start = 5, end = 9, frequency = 5)
plot(t)
```

***

For other variables in the data frame  
*Solar Radiation*:  
```{r timeseries}
var = df$Solar.R
t <- ts(var, start = 5, end = 9, frequency = 5)
plot(t)
```

***

*Temperature*:  
**Note**: R function arguments can usually be identified by position.
```{r timeseries2}
t <- ts(df$Temp, start = 5, end = 9, frequency = 5)
plot(t)
```


## Annotation
>- A graph or chart is incomplete without title, labels, legends and other kinds of annotation.

>- We will briefly examine some of the simpler approaches that could be employed in basic R plotting.

## Titles
```{r main}
title <- "Wind Speed for New York, May - Sept. 1973"
hist(df$Wind, main = title)
```

## Subtitles
```{r sub, eval=FALSE}
sub <- "Source: NY State Dept. of Conservation & US National Weather Service"
hist(df$Wind, main = title, sub = sub)
```

***

```{r sub, echo=FALSE}
```

## Axis Labels
X-Axis
```{r xlab, eval=FALSE}
x.label <- "Wind Speed (mph)"
hist(df$Wind, main = title, sub = sub, xlab = x.label)
```

***

```{r xlab, echo=FALSE}
```

***

Y-Axis
```{r ylab, eval=FALSE}
y.label <- "No. of observations"
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label
)
```

***

```{r ylab, echo=FALSE}
```

## Additional Labels
```{r labels, eval=FALSE}
labels = LETTERS[1:11]
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels      # earlier counted 11 bars
)      
```

***

```{r labels, echo=FALSE}
```

## Additional Text
Explanatory notes can be added to the plot
```{r mtext, eval=FALSE}
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels
) 
mtext("The data are approximately normally distributed.", side = 4)
```

***

```{r mtext, echo=FALSE}
```


## Customisation
>- We can customise our graphs and charts.
>- R gives the user several possibilities.
>- Limited only by imagination and skill.

## Axis limits
```{r ylim, eval=FALSE}
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels,
  ylim = c(0, 40)
)
```

***

```{r ylim, echo=FALSE}
```

***

```{r xlim, eval=FALSE}
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels,
  ylim = c(0, 40),
  xlim = c(0, 25)
)
```

***

```{r xlim, echo=FALSE}
```

## Colour
Fill:
```{r fill, eval=FALSE}
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels,
  ylim = c(0, 40),
  xlim = c(0, 25),
  col = "cyan"
)
```

***

```{r fill, echo=FALSE}
```

***
Line:
```{r line, eval=FALSE}
hist(
  df$Wind,
  main = title,
  sub = sub,
  xlab = x.label,
  ylab = y.label,
  labels = labels,
  ylim = c(0, 40),
  xlim = c(0, 25),
  col = "cyan",
  border = "red"
)
```

***

```{r line, echo=FALSE}
```


# Plotting Bivariate Data

*** 

```{r transform2, echo=FALSE}
df$Heat <- cut(df$Temp,
               breaks = c(55, 70, 85, 100),
               labels = c("Cold", "Normal", "Hot"),
               include.lowest = TRUE,
               ordered_result = TRUE)
```

```{r showcase, echo=FALSE, warning=FALSE}
# layout(matrix(c(rep(1, 5), 2, 2), 1, 7))
heatTable <- table(df$Heat, df$Mth.Name)
catColor <- c("lightblue", "brown", "red")

barplot(heatTable,
        main = "Ambient Cold/Heat in New York City, May - Sept. 1973",
        legend = FALSE,
        axes = TRUE,
        col = catColor,
        beside = TRUE,
        ylab = "No. of Days",
        ylim = c(0, 30))
# axis(2)

# ## Plot legend separately on account of space
# plot.new()
# legend("left", legend = levels(df$Heat), fill = catColor)
# layout(matrix(1))
```

## Data transformation
First, we will modify our data a bit by adding a column of temperature categories.

```{r}
head(df$Temp)
```

```{r transform2, eval=FALSE}
```
```{r out_heat_cat}
head(df$Heat, 10L)
```

***
`table()` gives the count of each discrete variable.
```{r table-heat}
table(df$Heat)
```

***
With 2 variables, we can make a 2 X 2 contingency table
```{r bivar-heat}
table(df$Heat, df$Mth.Name)
```

***

The table forms the basis for our bar chart
```{r plain_plot}
barplot(heatTable)
```

Default call on `barplot`. Use `?` to see all the arguments.

***

Change from the default stacked bar chart to a grouped bar chart
```{r}
barplot(heatTable, beside = TRUE)
```

***
Let's fix the y-axis

```{r fix axis}
barplot(heatTable, beside = TRUE, ylim = c(0, 30))
```

***

Our chart could do with some colour!
```{r}
barplot(heatTable, beside = TRUE, ylim = c(0, 30),
        col = c("lightblue", "brown", "red"))
```

***

What do the colours represent? We need a legend...

```{r}
barplot(heatTable, beside = TRUE, ylim = c(0, 30),
        col = c("lightblue", "brown", "red"), legend = TRUE)
```

***

Finally, we need to give this chart a title and a y-axis label.
```{r add title}
barplot(heatTable, beside = TRUE, ylim = c(0, 30),
        col = c("lightblue", "brown", "red"), legend = TRUE, 
        main = "Ambient Cold/Heat in New York City, May - Sept. 1973",
        ylab = "No. of Days")
```

# More fun plotting Air Quality Data

## Scatter plots with fitted line
```{r wnd-tmp plot, eval=FALSE}
plot(
  df$Wind,
  df$Temp,
  main = "Wind vs. Temperature",
  ylab = "Temperature (Fahrenheit)",
  xlab = "Wind (mph)",
  pch = 16,
  col = "blue"
)
mod <- lm(Temp ~ Wind, data = df)
coef <- mod$coefficients
abline(coef = coef, col = "red", lwd = 2)
legend(
  "bottomleft",
  legend = paste(names(coef), round(coef, 2)),
  bg = "gray",
  cex = .7
)
grid(col = "darkgrey")
```

***

```{r wnd-tmp plot, echo=FALSE}
```

***

```{r wnd-oz plot, eval=FALSE}
plot(
  df$Wind,
  df$Ozone,
  main = "Wind vs. Ozone",
  ylab = "Ozone (ppb)",
  xlab = "Wind (mph)",
  pch = 16,
  col = "red"
)
mod1 <- lm(Ozone ~ Wind, data = df)
coef1 <- mod1$coefficients
abline(coef = coef1, col = "blue", lwd = 2)
legend(
  "topright",
  legend = paste(names(coef1), round(coef1, 2)),
  bg = "gray",
  cex = .7
)
grid(col = "darkgrey")
```

***

```{r wnd-oz plot, echo=FALSE}
```

***

```{r end scatter, echo=FALSE}
cols <- c("blue", "green", "red", "black", "cyan")
plot(df$Ozone,
     df$Temp,
     main = "Ozone vs. Temperature",
     ylab = expression({"Temperature ("^0 + "F"}),
     xlab = "Ozone (ppb)",
     pch = 2,
     col = sapply(df$Mth.Name, function(s)
       switch(as.character(s),
              May = cols[1],
              June = cols[2],
              July = cols[3],
              August = cols[4],
              September = cols[5])))

legend("bottomright", legend = levels(df$Mth.Name), fill = cols, cex = .7)

grid()
```

***

Study the code later...

```{r end scatter, eval=FALSE}
```

# In conclusion

***

>- Indeed, R's graphics capabilities are what makes it to shine.
>- We have only scratched the surface
>- The key to learning to plot in R is practice and a lot of experimentation.
>- [Click here](https://www.datacamp.com/community/tutorials/15-questions-about-r-plots) to visit a very useful blog post on R plots

# Fin
