---
title: "Probability Distributions"
output:
  ioslides_presentation: default
  beamer_presentation:
    theme: metropolis
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Probabiility Functions
```{r}
nomnoml::nomnoml(
  "[Probability Mass Function] discrete variables ->[Cumulative Probability Function]
   [Probability Density Function] continuous variables ->[Cumulative Probability Function]"
)

```

## R Paradigm for Probability Distributions
Some of the key distributions:

>- binomial
>- Poisson
>- hypergeometric
>- exponential
>- normal
>- $\chi^2$
>- Student's t

## R Paradigm for Probability Distributions
- Family of functions; members identified by prefix
- 4 common prefixes
  + **d*** - density (or mass) function
  + **p*** - cumulative probability function
  + **q*** - inverse probability i.e. quantile
  + **r*** - random value generators

## Example: Binomial Distribution
```{r, eval = Sys.info()["nodename"] == "SA-DG"}
shiny::shinyAppDir("~/Documents/5-Personal/Study/datascience/R/r-sandbox/stats/binom/shiny-binomial")
```

