---
title: "Training on the naijR Package"
author: "DevSolutions Ltd."
date: "31/01/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(naijR)
```


Introduction

This inaugural training was held on `r format(Sys.Date(), "%A %d %B, %Y")`. It was an online event hosted via Zoom.

The package naijR was developed to provide much-needed help to researchers working with data related to Nigeria. The package is hosted on CRAN. To show the map of Nigeria, use the function map_ng(). The package website is found here.

This is a drawing of the simple map of Nigeria showing its constituent States.

```{r fct-map, warning=FALSE, fig.cap="Map of the FCT"}
map_ng(show.text = TRUE)
``` 

To install the package, use this code:

```{r, eval=FALSE}
install.packages('naijR')
```

Main features:

Innovative
Robust
Well tested
Informative maps
Helpful tools for data cleaning

Testimonials

"I am glad to have found this package. It has stimulated my interest in R programming as a whole".

"It is a very useful addition to data science tooling as a whole".

For more information, contact info@dev-solu.com.
