---
title: "Training on the naijR Package"
author: "DevSolutions Ltd."
date: "31/10/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(naijR)
```


# Introduction

This inaugural training was held on `r format(Sys.Date(), "%A %d %B, %Y")`. It was an online event hosted via Zoom.

The package *naijR* was developed to provide much-needed help to researchers working with data related to Nigeria. The package is hosted on CRAN. To show the map of Nigeria, use the function `map_ng()`. The package website is found [here](https://brovic.github.io/naijR/).

This is a drawing of the simple map of Nigeria showing its constituent States.

```{r fct-map, warning=FALSE, fig.cap="Map of Nigeria"}
map_ng(show.text = TRUE)
``` 

To install the package, use this code:

```{r, eval=FALSE}
install.packages('naijR')
```

Main features:

- Innovative
- Robust
- Well tested
- Informative maps
- Helpful tools for data cleaning

## Testimonials

> "I am glad to have found this package. It has stimulated my interest in R programming as a whole".

> "It is a very useful addition to data science tooling as a whole".

For more information, contact <mailto:info@dev-solu.com>.
