---
title: "Reproducible Data Science with R Markdown"
author: ""
date: "`r format(Sys.Date(), '%e %B, %Y')`"
output:
  ioslides_presentation: default
  powerpoint_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Motivation

--- 

Data analysis reports involve data, visualisations and narratives. Imagine this scenario:

- You are writing a report that also has data and results in it
  + mathematical output
  + tables
  + charts
- Your charts are beautiful. You made them in SPSS, Excel or R. You save and insert into Word.
- You realise you made a mistake with the data. You have to remake the charts, save them and insert into Word.
- You report is large and you have up to 30 tables. There was a mistake in the data entry that was discovered by a colleague. You now have to recreate 30 tables. 
- Poor you!

# What is R Markdown?

--- 

- R Markdown helps to prevent the aforementioned pain areas.
- Data analysis in itself can be a painful, gruelling process. Why add to it?
- Mistakes happen all the time and they are very costly in terms of time, energy and money.

## Some Background
- HTML
+ A markup language in which we have text files that have elements
+ The elements in the document are interpreted by applications called browsers.
+ The key aspect: interpreted by machine.

- Markdown:
Is also a markup language, but its name indicates that the markup is more managed.
The key is its readability.

## R
- R is a programming language and environment for statistical computing.
- In R Markdown, R code can be interspersed with text narrative to produce dynamic documents that respond to data driven events.

# The Components of an R Markdown document

---

There are 3 main pieces that fit together in an R Markdown document:

1. The YAML Header
2. Embedded R Code
3. Plain text

## The YAML Header
- YAML is a serialization language that provides easy to read-write data representations.
- In RMD, this is mostly used to define metadata for global settings of the document.
- For more information on YAML, visit <https://yaml.org>.

## The YAML Header

```yaml
---
title: "My Big Report"
author: "John Doe"
date: "2021-01-30"
output: word_document
---
```

## Embedded R Code (Chunks & Inline Code)
- R Code can be used inside the document in the form of chunks and inline:
- A code chunk is delineated with 3 backticks (```). To make it an code chunk wherein expressions can be evaluated, a language engine is defined. For R, this is written as __{r}__.

```md
```{r my-label-1, echo = FALSE}`r ''`
x <- 42
print(x)
```
```
Code chunks can carry an optional label, as well as arguments known as **chunk options** for finer control of what we want to allow a chunk to do.

## Embedded R Code

R code can be evaluated right inside the body of the document i.e. among the text characters.

To do this, we can use the \`r\ \` and whatever is inside those elements will be interpreted as R code and evaluated. The returned value of the expression will be rendered in the final document at that very position.

For example, if `x` is the name for an object with the value *8*, \`r\ x + 2\` will be printed as **10** in the final document.

## Text

- Plain text characters can be written as in any other electronic document format.
- Markup is applied by *appropriately* using particular symbolic characters.
- When the document is compiled, the markup is interpreted in line with the chosen format.

## Text

There are many ways to carry out text formatting:

- Headings are created with the '#' character (up to 6 levels)
- Bold text is made by surrounding it with ** or __.
- Italics are created by surrounding text with * or _

There are many other ways to do formatting e.g. mathematical symbols (MathJax)

## Writing the Document

This is done easily with R Studio

- Start a new file, save with the extension *.Rmd*.
- use various menus/icons to:
  + Create code chunks
  + Run chunks to check output

## Compiling the Document
The process has several steps:

1. Running of all code chunks and conversion to regular Markdown
2. Conversion of Markdown to document i.e. HTML, Word, PDF using Pandoc.
3. For PDF, LaTex is required to create and intermediate '.tex' file.

## Compiling the Document
- In R Studio, click the 'Knit' button.
- Using R code:
  + Useful when compiling many documents. For example, we can use a loop to create several of such documents
  + `rmarkdown::render`

<!-- ## Collaboration: Role of Version Control -->
<!-- - To collaborate effectively, we need version control software. This is the role played by Git. -->
<!-- - Git takes snapshots of the history of changes made in a source file. -->
<!-- - Multiple collaborators can seamlessly make changes to a source file, especially when a good workflow is used. -->

# Demo